---
title: "A Practical Tutorial on Gene Set Functional Enrichment Analysis"
author: "Dong Li"
date: 'April 25, 2017'
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivation

Functional enrichment analysis for gene lists is a computational approach that determines whether an a priori defined set of genes are significantly associated with biological functions. It has become a fundamental tool in biomedical research, which bridges the gap between algorithmic output gene set and prior knowledge, further gains the biological insight. There are various softwares implementing functional enrichment analysis using different databases, here we explore some of them from a practical pointview. Before that, the conceptual part of functional enrichment analysis is briefly introduced as below.

# Introduction
Here is a nice review on enrichment analysis^[Bioinformatics enrichment tools: paths toward the comprehensive functional analysis of large gene lists. *Nucleic acids research*, 2009.]. According to the review, enrichment tools can be categorized into three classes

 - singular enrichment analysis (SEA)
 - gene set enrichment analysis (GSEA)
 - modular enrichment analysis (MEA)

The core part is SEA, the most traditional strategy for enrichment analysis. Given a gene list, which can be differentially expressed (DE) genes or a module from a gene network, the basic idea is to compute the enrichment P-value, i.e. number of genes in the list that hit a given biology class as compared to pure random chance. Several statistical methods can be applied, including hypergeometric, binomial, chi-square, and Fisher's exact test^[Ontological analysis of gene expression data: current tools, limitations, and open problems, *Bioinformatics*, 2005.]. The hypergeometric formulation is directly derived from the problem if given several genes present an enrichment. Suppose we have a gene list $L$ with length $n$, and we want to know whether $X$ genes of them are sampled from a specific GO category $G$ with length $M$. The background population size is $N$. We have the following 2x2 table:

|     | in category  | not in category | Total   |
|-----+--------------+-----------------+---------|
| class 1 (the candidate)  | $X$   | $n-X$      | $n$ |
| class 2 (the rest) | $M-X$ | $N-M-n+X$  | $N-n$ |
| Total    | $M$   | $N-M$      | $N$ |

According to the [hypergeometric distribution](https://en.wikipedia.org/wiki/Hypergeometric_distribution), the probability mass function (pmf) of random variable $X$ is given by
$$P(X=k)=\frac{{M\choose k}{N-M\choose n-k}}{{N\choose n}}$$
where ${N\choose n}=\frac{N!}{k!\,(N-k)!}$ is the Binomial coefficient. The result $P(X=k)$ is the **probability** of $X=k$.

In other words, if none of the genes in the candidate list comes from the given GO category, we have to exclude all the possibilities (from 1 to $n$), then the P-value of this Hypergeometric Test is given by
$$p=1-\sum_{k=1}^n\frac{{M\choose k}{N-M\choose n-k}}{{N\choose n}}$$
If the p-value $p$ is very low (less than the alpha level 0.05), we can say the candidate gene list $L$ is significantly enriched by given GO category $G$. The rational keeps the same in pathway enrichment analysis, in which the GO category is replaced with a known pathway.

[Bonferroni correction](https://en.wikipedia.org/wiki/Bonferroni_correction) p-value is used when multiple hypotheses are tested. The goal is to maintain the probability of false find any significant hypothesis at the alpha level^[GO::TermFinder—open source software for accessing Gene Ontology information and finding significantly enriched Gene Ontology terms associated with a list of genes. *Bioinformatics*, 2004.]. If we have $m$ hypotheses, Bonferroni correction p-value is simply alpha value divided by $m$. We may also need adjusted p-value by Benjamini-Hochberg (BH) method for correction^[Controlling the False Discovery Rate: A Practical and Powerful Approach to Multiple Testing. *Journal of the Royal Statistical Society. Series B (Methodological)*, 1995.].

# Enrichment analysis in practise
## Basic analysis
Take the Affymetrix Human Genome U95 Set annotation data (chip hgu95av2) for example. Assume we have a gene list $prbs$, the population list $univ$. Each entry in both vector represent a gene  Entrez-ID. 
```{r, message=FALSE, warning=FALSE}
## Fake up some data
library("hgu95av2.db")
library("annotate")

prbs <- ls(hgu95av2GO)[1:300]
## Only those with GO ids
hasGO <- sapply(mget(prbs, hgu95av2GO), function(ids)
if(!is.na(ids) && length(ids) > 1) TRUE else FALSE)
prbs <- prbs[hasGO]
prbs <- getEG(prbs, "hgu95av2")
## remove duplicates, but keep named vector
prbs <- prbs[!duplicated(prbs)]
## do the same for universe
univ <- ls(hgu95av2GO)[1:5000]
hasUnivGO <- sapply(mget(univ, hgu95av2GO), function(ids)
if(!is.na(ids) && length(ids) > 1) TRUE else FALSE)
univ <- univ[hasUnivGO]
univ <- unique(getEG(univ, "hgu95av2"))
head(prbs)
head(univ)
```

The basic question: are these genes more often in the GO category (BP = Biological process; CC = Cellular component; MF = Molecular function) than expected by chance? Hypergeometric in package *GOstats* is used like:
```{r}
library("GOstats")
p <- new("GOHyperGParams", geneIds=prbs, universeGeneIds=univ,
         ontology="BP", annotation="hgu95av2", conditional=TRUE)

## this part takes time...
hyp <- hyperGTest(p)
head(summary(hyp))
```
A another complete example comes from [Martin Morgan](http://www.bioconductor.org/help/course-materials/2009/SSCMay09/gsea).

## Integrative anaysis
Powerful web-based enrichment analysis tools such as DAVID, Enrichr, GeneMANIA, Metascape, all support uploading gene list and return the result page with colorful plots and nice figures. There are wide range of databases for annotation and complicated algorithms behind each of them. Instead of introducing the web-based functions in details, here we conduct a simple case study using a programatic approach, which is more efficient when there are hundreds of gene lists to deal with. Take [DAVID Web Service](https://david.ncifcrf.gov/content.jsp?file=WS.html) for example, the url-based DAVID API allows users to programatically acess to DAVID as in the website and save the result locally. The following scrip works with [*RDAVIDWebService*](https://bioconductor.org/packages/release/bioc/html/RDAVIDWebService.html) package.
```{r,eval=FALSE}
library("RDAVIDWebService")
Haspgenelist <- readLines("xxfilename")
# replace xxx with your registered email
david <- DAVIDWebService(email=xxx,url="https://david.ncifcrf.gov/webservice/services/DAVIDWebService.DAVIDWebServiceHttpSoap12Endpoint/")
result <- addList(david, Haspgenelist,idType="ENTREZ_GENE_ID",
                  listName="genelist",listType="Gene")
setAnnotationCategories(david, c("GOTERM_BP_ALL","GOTERM_MF_ALL", "GOTERM_CC_ALL"))
termCluster <- getClusterReport(david, type="Term")
if (length(row.names(summary(termCluster))) > 0){
	getClusterReportFile(david,type="Term",
	                     "termClusterReport_dmel.tab")
	sample.dataframe <- read.csv(fileName,header=F,sep='\t')
	
  for (funclust in 1:length(row.names(summary(termCluster)))) {
    pdf(file=paste('FunctionalCluster_dmel_',funclust,'.pdf',
                 sep=""),compress=FALSE)
    print(plot2D(termCluster, funclust))
    dev.off()
	
    davidGODag <- DAVIDGODag(members(termCluster)[[funclust]],
                           pvalueCutoff=0.1, "CC")
    if (numNodes(goDag(davidGODag))>0){
        pdf(file=paste('GeneOntologyFunctionalCluster',
                     funclust,'.pdf',sep=""))
	    print(plotGOTermGraph(g=goDag(davidGODag),r=davidGODag,
	                        max.nchar=40, node.shape="ellipse"))
	    dev.off()
    }
  }

}

```
If there are several gene lists, each stored in a plain file, the above code can be wrapped with a `for` loop. And all final report plain files can be integrated into one file if necessary.

# Endnote
Functional enrichment analysis is probably only the first step of the complete explanation of the result gene lists, which needs more sound and detailed description about related functions. Ideally, the analysis can provide some useful insights and guidance for follow-up experiments. 

<!--But it would be the last step for a CS background guy. The question still remains as, where are the bounderies of a [interdisciplinary](https://en.wikipedia.org/wiki/Interdisciplinarity) research field located?-->

--
Last update: `r Sys.setlocale("LC_TIME", "English"); format(Sys.Date(), "%B %d, %Y")`